Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-04-04 optim.qmd
added pseudocode

---
title: Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliations: 
        - id: some-tech
          name: Some Institute of Technology
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - projection pursuit
  - optimization
  - jellyfish optimiser
  - data visualisation
  - high-dimensional data
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
---

@@

Below is the pseudo-code for this visualisation application.

\[Put the pseudo-code in, add specifics to this visualisation application\]

@@

An intuitive way to measure smoothness of a function would be to count how many continuous derivatives exist. To help define smoothness in our context, we can make use of the Sobolev spaces: functions $f$ in the Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$ continuous. Smoothness would then be the highest $p$ such that the index function belongs to $W^{p,\infty}$. We can relax this and consider $W^{p,q}$ Sobolev spaces.

@@

-   $h_i$ is an neighbourhood area around $a$. We can insert different values of $h$ to approximate the above quantity and observe how this quantity changes.

>>>>

---
title: Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
   - \usepackage{algpseudocode}
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliations: 
        - id: some-tech
          name: Some Institute of Technology
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - projection pursuit
  - optimization
  - jellyfish optimiser
  - data visualisation
  - high-dimensional data
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
---

@@

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the 'best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial $i$ and `max.tries`). Occasionally, we might explore
completely new directions like a jellyfish might with ocean currents.
Then, we compare new potential projections to our current one. If
they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

::: {.callout-note icon="false"}

## Algorithm: Jellyfish Optimizer Pseudo Code

**Input**: `current_projections`, `index_function`, `tries`, `max_tries`

**Output**: `optimized_projection`

**Initialize** `best_jelly` as the projection with the best index value from `current_projections`, and 
`current_index` as the array of index values for each projection in `current_projections`

>

**for** each try in 1 to max_tries **do**

> Calculate $c_t$ based on the current try and max_tries

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>    
> > Define trend based on the best jelly and current projections
>     
> > Update each projection towards the trend using a random factor and orthonormalisation
>       
> **else**
>     
> > **if** a random number is greater than $1 - c_t$ **then**
> >       
> > > Slightly adjust each projection with a small random factor (Type A passive)
> >         
> > **else**
> >       
> > > For each projection, compare with a random jelly and adjust towards or away from it (Type B active)
>         
> Update the orientation of each projection to maintain consistency
>     
> Evaluate the new projections using the index function

>

> **if** any new projection is worse than the current, revert to the `current_projection` for that case
> 
> > Determine the projection with the best index value as the new best_jelly

>

> **if** the try is the last one, print the final best projection and **exit**

**return** the set of projections with the updated best_jelly as the optimized_projection

:::

<!-- Var names in the function and in the text needs to be consistent??? -->

@@

<!--
An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. 
To help define smoothness in our
context, we can make use of the Sobolev spaces: functions $f$ in the
Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$
continuous. Smoothness would then be the highest $p$ such that the index
function belongs to $W^{p,\infty}$. We can relax this and consider
$W^{p,q}$ Sobolev spaces.

@@

-   $h_i$ is an neighbourhood area around $a$. We can insert different
    values of $h$ to approximate the above quantity and observe how this
    quantity changes.
-->

If we evaluate the index function at some random points (like the random initialization of the jellyfish optimizer), then we can interpret these random index values as a random field, indexed by a space parameter: the random projection angle. This analogy suggests to use this random training sample to fit a spatial model, a simple one being a (spatial) Gaussian process. There exist R packages to fit the hyperparameters of a Gaussian process, for example ExaGeoStatR (https://github.com/ecrc/exageostatR), which is an R wrapper for the ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).

How can we define a measure of smoothness from this? A Gaussian process is fully defined by its mean function and covariance function. The way the covariance function is defined is where smoothness comes into play: if an index is very smooth, then two close projection angles should produce close index values (strong correlation); by contrast, if an index is not smooth, then two close projection angles might give very different index values (fast decay of correlations with respect to distance between angles).

Luckily, the most popular class of covariance functions, known as Matérn covariance functions (see https://arxiv.org/pdf/1708.02835.pdf on page 3) has one parameter, Theta3, which controls the decay of the covariance function, which in turns controls the smoothness of the random field. The ExaGeoStat paper calls it smoothness parameter: the higher Theta3, the smoother the random field. In our context, the higher Theta3, the smoother the index function.

This suggests the following approach:

- Evaluate the index function at some random projection angle values (our initial jellies),
- Use this sample to fit a Gaussian process with Matérn covariance using ExaGeoStatR,
- Use the fitted Theta3 parameter value as our definition of smoothness of the index function: the higher Theta3, the smoother the index.

### 2024-04-04 optim.tex
added pseudocode

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit},
  pdfauthor={Alice Anonymous; Bob Security; Cat Memes; Derek Zoolander},
  pdfkeywords={projection pursuit, optimization, jellyfish
optimiser, data visualisation, high-dimensional data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}

@@

\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}

@@

\section{Projection pursuit, index functions and optimisation {[}Di and
Sherry{]}}\label{sec-background}

@@

\section{The jellyfish optimiser and properties of PP indexes {[}Nicolas
and Jessica{]}}\label{sec-theory}

@@

Below is the pseudo-code for this visualisation application.

{[}Put the pseudo-code in, add specifics to this visualisation
application{]}

@@

\subsection{Smoothness}\label{smoothness}

An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. To help define smoothness in our
context, we can make use of the Sobolev spaces: functions \(f\) in the
Sobolev space \(W^{p,\infty}\) have all derivatives of order less than
\(p\) continuous. Smoothness would then be the highest \(p\) such that
the index function belongs to \(W^{p,\infty}\). We can relax this and
consider \(W^{p,q}\) Sobolev spaces.

Consider the following definition of partial derivative. Let \(U\) be an
open subset of \(\mathbf{R}^n\) and \(f: U \rightarrow \mathbf{R}\). The
partial derivative of function \(f\) at the point
\(\mathbf{a} = (a_1, \ldots, a_n) \in U\) with respective to the
\(i\)-th variable \(x_i\) is defined as

\[
\frac{\delta}{\delta x_i} f(\mathbf{a}) =  \lim_{h \rightarrow 0} \frac{f(a_1, \ldots, a_{i-1}, a_i +h, a_{i+1}, \ldots, a_n) - f(a_1 , \ldots, a_i, \ldots, a_n)}{h} = \lim_{h\rightarrow 0} \frac{f(\mathbf{a} + h\mathbf{e_i}) - f(\mathbf{a})}{h}.
\] where \(\mathbf{e_i}\) is the unit vector of the \(i\)-th variable
\(x_i\).

If the derivative is not well-defined, we propose to approximate the
derivative using the following expression:

\[\frac{1}{n_i}\sum_{i} \frac{\|f(\mathbf{a} + h_i\mathbf{e_i}) - f(\mathbf{a})\|^p}{h_i}\]

where \(p \in [0,1]\). The choice of \(p\) reflects the penalty
behaviour. For the application in this paper, we choose \(p = 1\).

\begin{itemize}
\tightlist
\item
  \(h_i\) is an neighbourhood area around \(a\). We can insert different
  values of \(h\) to approximate the above quantity and observe how this
  quantity changes.
\end{itemize}

\subsection{Squintability}\label{squintability}

@@

\subsection{Speed}\label{speed}

@@

\section{Application {[}Di and Sherry{]}}\label{sec-simulation}

@@

\subsection{Going beyond 10D}\label{going-beyond-10d}

@@

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

\end{figure}%

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-3-1.pdf}

@@

\end{figure}%

\subsection{On skewness and kurtosis
index}\label{on-skewness-and-kurtosis-index}

\subsection{Another data example}\label{another-data-example}

\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit},
  pdfauthor={Alice Anonymous; Bob Security; Cat Memes; Derek Zoolander},
  pdfkeywords={projection pursuit, optimization, jellyfish
optimiser, data visualisation, high-dimensional data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, frame hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction-nicolas-and-jessica}{%
\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}}

@@

\hypertarget{sec-background}{%
\section{Projection pursuit, index functions and optimisation {[}Di and
Sherry{]}}\label{sec-background}}

@@

\hypertarget{sec-theory}{%
\section{The jellyfish optimiser and properties of PP indexes {[}Nicolas
and Jessica{]}}\label{sec-theory}}

@@

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the `best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial \(i\) and \texttt{max.tries}). Occasionally, we might
explore completely new directions like a jellyfish might with ocean
currents. Then, we compare new potential projections to our current one.
If they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, opacitybacktitle=0.6, leftrule=.75mm, left=2mm, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, breakable, colbacktitle=quarto-callout-note-color!10!white, colback=white, rightrule=.15mm, opacityback=0, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm]

\textbf{Input}: \texttt{current\_projections}, \texttt{index\_function},
\texttt{tries}, \texttt{max\_tries}

\textbf{Output}: \texttt{optimized\_projection}

\textbf{Initialize} \texttt{best\_jelly} as the projection with the best
index value from \texttt{current\_projections}, and
\texttt{current\_index} as the array of index values for each projection
in \texttt{current\_projections}

\begin{quote}
\end{quote}

\textbf{for} each try in 1 to max\_tries \textbf{do}

\begin{quote}
Calculate \(c_t\) based on the current try and max\_tries
\end{quote}

\begin{quote}
\textbf{if} \(c_t\) is greater than or equal to \(0.5\) \textbf{then}

\begin{quote}
Define trend based on the best jelly and current projections
\end{quote}

\begin{quote}
Update each projection towards the trend using a random factor and
orthonormalisation
\end{quote}

\textbf{else}

\begin{quote}
\textbf{if} a random number is greater than \(1 - c_t\) \textbf{then}

\begin{quote}
Slightly adjust each projection with a small random factor (Type A
passive)
\end{quote}

\textbf{else}

\begin{quote}
For each projection, compare with a random jelly and adjust towards or
away from it (Type B active)
\end{quote}
\end{quote}

Update the orientation of each projection to maintain consistency

Evaluate the new projections using the index function
\end{quote}

\begin{quote}
\end{quote}

\begin{quote}
\textbf{if} any new projection is worse than the current, revert to the
\texttt{current\_projection} for that case

\begin{quote}
Determine the projection with the best index value as the new
best\_jelly
\end{quote}
\end{quote}

\begin{quote}
\end{quote}

\begin{quote}
\textbf{if} the try is the last one, print the final best projection and
\textbf{exit}
\end{quote}

\textbf{return} the set of projections with the updated best\_jelly as
the optimized\_projection

\end{tcolorbox}

@@

\hypertarget{smoothness}{%
\subsection{Smoothness}\label{smoothness}}

If we evaluate the index function at some random points (like the random
initialization of the jellyfish optimizer), then we can interpret these
random index values as a random field, indexed by a space parameter: the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process. There exist R packages to fit the hyperparameters of a
Gaussian process, for example ExaGeoStatR
(https://github.com/ecrc/exageostatR), which is an R wrapper for the
ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).

How can we define a measure of smoothness from this? A Gaussian process
is fully defined by its mean function and covariance function. The way
the covariance function is defined is where smoothness comes into play:
if an index is very smooth, then two close projection angles should
produce close index values (strong correlation); by contrast, if an
index is not smooth, then two close projection angles might give very
different index values (fast decay of correlations with respect to
distance between angles).

Luckily, the most popular class of covariance functions, known as Matérn
covariance functions (see https://arxiv.org/pdf/1708.02835.pdf on page
3) has one parameter, Theta3, which controls the decay of the covariance
function, which in turns controls the smoothness of the random field.
The ExaGeoStat paper calls it smoothness parameter: the higher Theta3,
the smoother the random field. In our context, the higher Theta3, the
smoother the index function.

This suggests the following approach:

\begin{itemize}
\tightlist
\item
  Evaluate the index function at some random projection angle values
  (our initial jellies),
\item
  Use this sample to fit a Gaussian process with Matérn covariance using
  ExaGeoStatR,
\item
  Use the fitted Theta3 parameter value as our definition of smoothness
  of the index function: the higher Theta3, the smoother the index.
\end{itemize}

\hypertarget{squintability}{%
\subsection{Squintability}\label{squintability}}

@@

\hypertarget{speed}{%
\subsection{Speed}\label{speed}}

@@

\hypertarget{sec-simulation}{%
\section{Application {[}Di and Sherry{]}}\label{sec-simulation}}

@@

\hypertarget{going-beyond-10d}{%
\subsection{Going beyond 10D}\label{going-beyond-10d}}

@@

\begin{figure}

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-4-1.pdf}

@@

\end{figure}

\begin{figure}

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-6-1.pdf}

@@

\end{figure}

\hypertarget{on-skewness-and-kurtosis-index}{%
\subsection{On skewness and kurtosis
index}\label{on-skewness-and-kurtosis-index}}

\hypertarget{another-data-example}{%
\subsection{Another data example}\label{another-data-example}}

\hypertarget{sec-conclusion}{%
\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}}
